---
title: "Take-Home Assignment"
subtitle: "Dose Response Analysis" 
author: "Yubing Wan, Ph.D."
date: "2024-01-20"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction 
- Four Parameter Logistic (4PL) Regression Model 
- Computing Programs 
  - R 
  - Python 

## Equation and Parameters of the 4PL Model

$$
\begin{aligned}
& f(X, HillSlope, top, bottom, EC_{50}) = \\
& \\
& bottom + \frac{top - bottom}{1 + 10^{(LogEC_{50} - Log(X)) * HillSlope}}
\end{aligned}
$$

## Parameter Estimation and Curve Fitting in R 

```{r, results = 'hide', message = FALSE}
if (!require("tidyverse", character.only = TRUE)) {
    install.packages("tidyverse", dependencies = TRUE)
    library("tidyverse", character.only = TRUE)
}

if (!require("drc", character.only = TRUE)) {
    install.packages("drc", dependencies = TRUE)
    library("drc", character.only = TRUE)
}

if (!require("broom", character.only = TRUE)) {
    install.packages("broom", dependencies = TRUE)
    library("broom", character.only = TRUE)
}

if (!require("reticulate", character.only = TRUE)) {
    install.packages("reticulate", dependencies = TRUE)
    library("reticulate", character.only = TRUE)
}

theme_set(theme_bw())
source(file.path(str_remove(getwd(), "/inst/rmd"), "R/funcs.R"))

## Define demo datasets
data_demo_stim <- data.frame(
  dose = c(1e-9 , 1e-8 , 1e-7 , 1e-6 , 1e-5 , 1e-4 , 1e-3 , 1e-2), 
  response = c(0, 3, 8, 19, 31, 38, 40, 41))
  
data_demo_inhi <- data.frame(
  dose = c(1e-9 , 1e-8 , 1e-7 , 1e-6 , 1e-5 , 1e-4 , 1e-3 , 1e-2), 
  response = rev(c(0, 3, 8, 19, 31, 38, 40, 41)))
```

### Inhibition 

```{r, results = 'hide'}
m1 <- fit_4pl_model(data_demo_inhi)
m2 <- fit_4pl_model(data_demo_inhi, constr_top = 35, algorithm = "port")

par(mfrow = c(1, 2))
plot_dose_response_curve(m1, data_demo_inhi, main = "Without Constriaints")
plot_dose_response_curve(m2, data_demo_inhi, main = "With Top Constriaint")
```

## 
### Stimulation

```{r, results = 'hide'}
m1 <- fit_4pl_model(data_demo_stim)
m2 <- fit_4pl_model(data_demo_stim, constr_bottom = 10, algorithm = "port")
m3 <- fit_4pl_model(data_demo_stim, constr_top = 35, algorithm = "port")
m4 <- fit_4pl_model(data_demo_stim, constr_bottom = 10, constr_top = 35, algorithm = "port")

par(mfrow = c(2, 2))
plot_dose_response_curve(m1, data_demo_stim, main = "Without Constriaints")
plot_dose_response_curve(m2, data_demo_stim, main = "Bottom Constriaint")
plot_dose_response_curve(m3, data_demo_stim, main = "Top Constriaint")
plot_dose_response_curve(m4, data_demo_stim, main = "Both Constriaints")
```

## Simulation of Realistic Datasets 

## Inhibition
```{r}
set.seed(123) # For reproducibility
inhibition_data <- simulate_dose_response_data(type = "inhibition")
plot(inhibition_data$dose, inhibition_data$response, col = "blue", main = "Simulated Inhibition Data")
fit <- fit_4pl_model(inhibition_data, log50_0 = log10(50), h_0=1, top_0 = 100)
y <- fit$bottom + (fit$top - fit$bottom)/(1 + 10^((fit$logEC50 - log10(inhibition_data$dose)) * fit$h)) # Y data
lines(inhibition_data$dose, y, col="blue", lwd=2) # plot the curve
```

## Stimulation
```{r}
stimulation_data <- simulate_dose_response_data(type = "stimulation")
# fit_drm <- dose_response_analysis(stimulation_data, response_type = "stimulation", constrained = T)
# fit_drm$coefficients
# par(mfrow = c(2, 1))
# plot(fit_stimulation, main = "crd")
plot(stimulation_data$dose, stimulation_data$response, col = "red", main = "Simulated Stimulation Data")
fit <- fit_4pl_model(stimulation_data, log50_0 = log10(50), h_0=1, top_0 = 100)
y <- fit$bottom + (fit$top - fit$bottom)/(1 + 10^((fit$logEC50 - log10(inhibition_data$dose)) * fit$h)) # Y data
lines(stimulation_data$dose, y, col="red", lwd=2) # plot the curve
```


## Cross-Software Validation 

- Parameter Estimation and Curve Fitting in Python 

```{python}
import subprocess
import pkg_resources
required_packages = ["numpy", "scipy", "matplotlib"] 
installed_packages = {pkg.key for pkg in pkg_resources.working_set}
missing_packages = [pkg for pkg in required_packages if pkg not in installed_packages]
import numpy as np
import matplotlib.pyplot as plt
from scipy.optimize import curve_fit

if missing_packages:
    subprocess.check_call(["python", "-m", "pip", "install", *missing_packages])


def four_parameter_logistic(x, top, bottom, ec50, hill_slope):
    return bottom + (top - bottom) / (1 + (x / ec50)**(-hill_slope))

def fit_4pl_model(x_data, y_data, initial_guess=(100, 0, 50, 1)):
    popt, _ = curve_fit(four_parameter_logistic, x_data, y_data, p0=initial_guess, maxfev=10000)
    return popt
  
def calculate_fitted_response(dose, popt):
    top, bottom, ec50, hill_slope = popt
    fitted_response = four_parameter_logistic(dose, top, bottom, ec50, hill_slope)
    return fitted_response

def simulate_dose_response(n=100, response_type='inhibition', ec50=50, 
                           hill_coeff=1, max_response=100, min_response=0, noise_level=0.05):
    dose = np.linspace(min_response, max_response, n)

    if response_type == 'inhibition':
        response = max_response / (1 + (dose/ec50)**hill_coeff)
    elif response_type == 'stimulation':
        response = min_response + (max_response - min_response) * (dose**hill_coeff / (ec50**hill_coeff + dose**hill_coeff))
    else:
        raise ValueError("response_type must be 'inhibition' or 'stimulation'")

    # Add random noise to simulate experimental variability
    response += np.random.normal(0, noise_level * max_response, n)

    return dose, response
    
def plot_dose_response_with_fit(dose, response, fitted_response, title="Dose-Response Curve", color = "blue"):
    plt.figure(figsize=(8, 5))
    plt.scatter(dose, response, color=color, label='Observed Data')  # Plot observed data points
    plt.plot(dose, fitted_response, color=color, label='Fitted Curve')  # Plot fitted curve
    plt.title(title)
    plt.xlabel("Dose")
    plt.ylabel("Response")
    plt.legend()
    plt.grid(True)
    plt.show()
```

## Inhibition 

```{python, message = F, results = "hide"}
dose, response = simulate_dose_response(response_type='inhibition')
popt = fit_4pl_model(dose, response)
fitted_response = calculate_fitted_response(dose, popt)
plot_dose_response_with_fit(dose, response, fitted_response, title="Inhibited Dose-Response Curve")
```

## Stimulation
```{python} 
dose, response = simulate_dose_response(response_type='stimulation')
popt = fit_4pl_model(dose, response)
fitted_response = calculate_fitted_response(dose, popt)
plot_dose_response_with_fit(dose, response, fitted_response, title="Simulated Dose-Response Curve", color = "red")
```


## Discussion 
